Legacy Data Confounds Modern Studies

### 2018-04-11 Draft3.tex
more words

\newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
\newcommand{\latcomment}[1]{\textcolor{blue}{LAT: #1}}
\newcommand{\todo}[1]{\textcolor{blue}{TODO: #1}}

@@

\maketitle
\tableofcontents
\clearpage
			\section{Introduction}
The last 5 years has seen an increase in number of individuals genotyped through private companies as well as medical research cohorts. Genotype data is often imputed to increase power for association studies in a cost effective way. 
This process is done using a reference database comprised of whole genome sequence data from other individuals in reference populations. 
We often combine these reference databases to increase the sample size and in turn, the accuracy of imputation. Combining data can be an issue when the quality of the data produced can vary between sequencing technologies and even sequencing centres.  
The errors in one dataset don't disappear when they are combined with another higher quality one. 
Many of the legacy datasets have been known to contain a higher rate of false positives than their more recent counterparts. 
Is it time to retire legacy datasets?

In investigating a mutational signature identified in a Japanese cohort by Harris et. al (2017), we noticed that the individuals responsible for this signal all had lower average quality scores compared to the other Japanese individuals. 
To further investigate the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS). 
We discovered 000 variants that were associated to low mapping quality in \todo{000} individuals from 000 of the 26 populations included in the 1000 genomes dataset. 
One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.

@@

@@

@@

@@

\afterpage{
\begin{figure}
\includegraphics[width=\hsize]{figures/MapQualOverTime.jpg}

\caption{A) The average mapping quality of each individual per population included in the 1000 Genomes sequencing project. The average mapping quality per individual was obtained from the raw alignment index data for each sample. The X axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual. B) Same data as in A) except the X axis is now sorted by sequencing date, and the colors indicate the sequencing centers that produced the data for each individual.}
\label{Figure1}
\end{figure}
\clearpage
}

\afterpage{
\begin{landscape}
\begin{figure}[]
\centering
\includegraphics[width=23cm,keepaspectratio, right]{figures/GWAS_Qual_GenomeWide_JPT.jpg}
\caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project. This GWAS included 104 individuals, 000 of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project. The same analysis was performed independently for each of the populations in the 1000 Genomes Project. }
 \label{Figure2}
\end{figure}
\end{landscape}

\clearpage

  }

\afterpage{
\begin{figure}[h]
\centering
\includegraphics[width=\hsize,keepaspectratio, right]{figures/SNPOverlap6.jpg}

\caption{Overlap of SNPs identified independently to be associated with quality. Not surprisingly, the populations that have the most low-quality individuals also have the most low-quality variants. What is interesting here, is that the same variants identified as being low quality independently in each population are found in other populations. }
  \label{Figure3}
\end{figure}
\clearpage
}

>>>>

\newcommand{\sgcomment}[1]{\textcolor{blue}{SG: #1}}
\newcommand{\lat}[1]{\textcolor{blue}{LAT: #1}}
\newcommand{\todo}[1]{\textcolor{blue}{TODO: #1}}

@@

\maketitle
\tableofcontents
\clearpage
			\section{Introduction}
\subsection{Larger Sample Sizes}			
The last 5 years has seen an increase in the number of individuals genotyped through private companies or medical research cohorts. 
\lat{not a fan of this sentence}
Larger sample sizes allow researchers to identify finer resolution statistically significant differences between groups of individuals. 
Mutations associated to a disease can be identified by comparing individuals afflicted by a disease to healthy individuals. 
The mutations more common in one group might be associated to the disease.
Demonstrating that these mutations are biologically relevant to the disease can be difficult.
Certain factors like age, sex or ancestral origin might be correlated to the trait in question and could potentially be mistakenly associated to a disease. 
Especially with the increasing size of cohorts, spurious associations are increasingly becoming an issue. 
For this reason, careful consideration must be taken when including individuals from different ancestral origins in these association studies. 
Benign mutations at high frequencies in one populations might be exceedingly rare in another population. 
Therefore, population wide differences in mutations must be included as covariates to avoid spurious associations.

\subsection{Imputation}
 improvements in the identification of rare genetic diseases as well as the risk prediction to certain types of cancer and diseases. 
Genotype data is often imputed to increase power for association studies in a cost effective way. 
This process is done using a reference database comprised of whole genome sequence data from other individuals in reference populations. 
We often combine these reference databases to increase the sample size and in turn, the accuracy of imputation. 
Combining data can be an issue when the quality of the data produced can vary between sequencing technologies and even sequencing centres.  
The errors in one dataset don't disappear when they are combined with another higher quality one. 
Many of the legacy datasets have been known to contain a higher rate of false positives than their more recent counterparts. 
Is it time to retire legacy datasets?

\lat{sentence about harris} 
A genome wide mutational signature can be measured by taking the sum of all the different types of mutations in an individual. 
The signatures from individuals from the same population will have a tendency to be more similar due to shared ancestry.
Harris et al have recently proposed that there are patterns of mutations that differ across human populations.
They argued this by showing that relative difference in mutational types between and even within continental populations.  
In investigating a mutational signature identified in a Japanese cohort by Harris et. al (2017), we noticed that the individuals responsible for this signal all had lower average quality scores compared to the other Japanese individuals. 
To further investigate the extent to which low quality individuals were associated to spurious mutations, we then performed a genome-wide association study (GWAS). 
We discovered \todo{000} variants that were associated to low mapping quality in \todo{000} individuals from \todo{000} of the 26 populations included in the 1000 genomes dataset. 
One third of these mutations were included in imputation tests. Population specific mutational signatures were also clarified by removing these spurious mutations.

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{MapQualOverTime.jpg}

\caption{A) The average mapping quality of each individual per population included in the 1000 Genomes sequencing project. The average mapping quality per individual was obtained from the raw alignment index data for each sample. The X axis is ranked by populations with the lease to the most variance, followed by average mapping quality per individual. B) Same data as in A) except the X axis is now sorted by sequencing date, and the colors indicate the sequencing centers that produced the data for each individual.}
\label{Figure1}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{GWAS_Qual_GenomeWide_JPT.jpg}
\caption{Genome wide association of the average quality of mapped bases for the Japanese individuals included in the 1000 Genomes Project. This GWAS included 104 individuals, 000 of them were sequenced in phase 1 while the rest were sequenced in phase 3 of the 1000 Genomes Project. The same analysis was performed independently for each of the populations in the 1000 Genomes Project. }
 \label{Figure2}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\hsize,keepaspectratio]{SNPOverlap6.jpg}

\caption{Overlap of SNPs identified independently to be associated with quality. Not surprisingly, the populations that have the most low-quality individuals also have the most low-quality variants. What is interesting here, is that the same variants identified as being low quality independently in each population are found in other populations. }
  \label{Figure3}
\end{figure}

@@
